---
title: "COVID Excess Death "
author: Viguen Poghosyan, Davit Sargsyan
version: 1.0
output:
  html_notebook:
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: yes
    code_folding: hide
---

Date: `r date()`

Meeting 10/28/22
Javier:
1. use "decompose" function (Fourier transform)
2. find top 3 frequencies (for 12-month frequency)
3. remove 12-months signal
4. run 'decompose' again for longer frequency (24 or 36 month?)  
2. Repeat seasonality analysis for each of the 20 states and for each cause of death.  
3. How to make this analysis for all 20 states in a single model (similar to random effects model idea)?  
4. Use new data from Nuria (US-wide, not separated by states).    
5. F-Test residuals from the 12month and 13 month (seasonality of residuals). Other ways to compare? bootstrap?  
6. Try to get daily temperatures. What part of months are the peaks  - beginning, middle or end? are the peaks between 2 months?  
7. Using lockdown, mask and vaccination data, model residuals of the forecast (COVID years). Might be lag. Talk to Yajie.  
8. Weighted PCA? Talk to Yajie.  
9. For PCA, transform log(1+ a%) - log(1 - b%). Use DMR::trans2 function.  

# Setup
```{r setup}
# install.packages("C:/git_local/DNAMR_1.2.tar.gz",
#                  repos = NULL,
#                  type = "source")
## NOTE: needs dependencies: moments

require(data.table)
require(ggplot2)
require(nlme)
# require(DNAMR)
```

# Data
```{r}
dt1 <- data.table(read.csv("data/JOINT_HistRec_till23_state_month_rev7_Tableau.csv"))

dt2 <- dt1[, .(Year = Year,
               Date = as.Date(paste0(Year,
                                     "/",
                                     Month,
                                     "/01"),
                              format = "%Y/%m/%d"),
               State = factor(State),
               Population = as.integer(Population),
               Deaths = as.integer(Deaths),
               daysmonth = daysmonth,
               DeathDaily = Deaths/daysmonth,
               COVID_deaths = COVID_deaths,
               COVIDDeathDaily = COVID_deaths/daysmonth)]
summary(dt2)
```
## Remove records after 023-06-01 and all COVID deaths
NOTE: Death data looks unreliable after April 2023. Remove May-June.  
  
```{r}
dt2 <- droplevels(dt2[Date < "2023-05-01", ]) 
```

## State populations
```{r}
tmp <- unique(dt2[, c("Date",
                      "State",
                      "Population")])
range(tmp$Date)

tmp[, pop2015 := Population[Date == "2015-01-01"],
    by = State]
tmp[, pop2015pct := 100*Population/pop2015]
```

```{r,fig.height=10,fig.width=15}
p1 <- ggplot(tmp,
             aes(x = Date,
                 y = Population)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(tmp$Date),
                                       to = max(tmp$Date),
                                       by = 365.25)) +
  ggtitle("Population") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

# tiff(filename = "tmp/population.tiff",
#      height = 10,
#      width = 15,
#      units = 'in',
#      res = 300,
#      compression = "lzw+p")
# print(p1)
# graphics.off()

p1
```

```{r,fig.height=10,fig.width=15}
p2 <- ggplot(tmp,
             aes(x = Date,
                 y = pop2015pct)) +
  facet_wrap(~ State) +
  geom_hline(yintercept = 100,
             linetype = "dashed",
             color = "red") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(tmp$Date),
                                       to = max(tmp$Date),
                                       by = 365.25)) +
  ggtitle("Population (% January 2015)") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

# tiff(filename = "tmp/population_pct_jan010.tiff",
#      height = 10,
#      width = 15,
#      units = 'in',
#      res = 300,
#      compression = "lzw+p")
# print(p2)
# graphics.off()

p2
```

## Daily death
```{r,fig.height=10,fig.width=15}
p1 <- ggplot(dt2,
             aes(x = Date,
                 y = DeathDaily)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                       to = max(dt2$Date),
                                       by = 365.25)) +
  ggtitle("Daily death") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

# tiff(filename = paste0("tmp/death_daily.tiff"),
#      height = 10,
#      width = 15,
#      units = 'in',
#      res = 300,
#      compression = "lzw+p")
# print(p1)
# graphics.off()

p1
```

## COVID-19 only
```{r,fig.height=10,fig.width=15}
p1 <- ggplot(dt2,
             aes(x = Date,
                 y = COVIDDeathDaily)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = as.Date("2020-03-01"),
                                       to = as.Date("2022-07-01"),
                                       by = 365.25/4)) +
  ggtitle("COVID-19 (U07.1) death daily average per month ('DeathDaily')") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```

## Calculate the rate (no more zeros!)
```{r}
# Rate = 10^5*(dt1$Deaths + 1)/(daysmonth*(Population + 1))
dt2[, Rate := 10^5*DeathDaily/Population]
```

## Daily death rate
```{r,fig.height=10,fig.width=15}
p1 <- ggplot(dt2,
             aes(x = Date,
                 y = Rate)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = as.Date("2015-01-01"),
                                       to = as.Date("2023-05-01"),
                                       by = 365.25/4)) +
  ggtitle("All-Cause Daily Death Rate") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```

# All states together
```{r}
dt_all <- dt2[, .(all_deaths = sum(Deaths),
                  all_pop = sum(Population),
                  all_covid_deaths = sum(COVID_deaths)),
              by = list(Date,
                        daysmonth)]
dt_all[, all_deaths_daily := all_deaths/daysmonth]
dt_all[, all_covid_deaths_daily := all_covid_deaths/daysmonth]
```

## Daily death all states
```{r,fig.height=5,fig.width=8}
p1 <- ggplot(dt_all,
             aes(x = Date,
                 y = all_deaths_daily)) +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                       to = max(dt2$Date),
                                       by = 365.25)) +
  ggtitle("All States Daily All Cause Death") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```

## Daily COVID-19 death all states
```{r,fig.height=5,fig.width=8}
p1 <- ggplot(dt_all,
             aes(x = Date,
                 y = all_covid_deaths_daily)) +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                       to = max(dt2$Date),
                                       by = 365.25)) +
  ggtitle("All States Daily COVID-19 Death") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```

# Training/Testing/COVID
```{r}
dt2$set <- "trn"

dt2$set[dt2$Date >= "2019-03-01" &
          dt2$Date < "2020-03-01"] <- "tst"

dt2$set[dt2$Date >= "2020-03-01"] <- "cvd"
```

# Modeling
```{r}
dt2$dt <- as.numeric(dt2$Date)
dt2$dt_sqr <- dt2$dt^2
```

## LM
```{r}
m1 <- lm(Rate ~ (dt + dt_sqr)*State,
         data = dt2[set == "trn"])

dt2$prd1 <- predict(m1,
                    newdata = dt2)
```
  
## Plot LM
```{r,fig.height=10,fig.width=15}
p1 <- ggplot(dt2,
             aes(x = Date,
                 y = Rate,
                 group = State)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_ribbon(aes(xmin = min(dt2$Date[dt2$set == "tst"]),
                  xmax = max(dt2$Date[dt2$set == "tst"]),
                  ymin = -Inf,
                  ymax = Inf),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = prd1),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                       to = max(dt2$Date),
                                       by = 365.25)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
p1
```

## NLME
```{r}
m3 <- nlme(Rate ~ b0 + b1*dt + b2*dt^2 + b3*sin(b4 + b5*dt),
           data = dt2[set == "trn"],
           random = b0 + b1 + b2 + b3 + b4 + b5 ~ 1 | State,
           fixed = b0 + b1 + b2 + b3 + b4 + b5 ~ 1,
           start = list(fixed = c(b0 = 2.5, # linear intercept
                                  b1 = 1, # linear slope
                                  b2 = 1, # square term, i.e. non-linearity of the trend
                                  b3 = 10000, # wave magnitude
                                  b4 = 1, # sin shift to the left
                                  b5 = 0.018)),# wave length(> 1 shortens, < 1 elongates)
           control = nlmeControl(maxIter = 500)) 
m3

dt2$prd3 <- predict(m3,
                    newdata = dt2)

# Save coefficients
betas <- m3$coefficients$fixed

# Save Mean Absolute Percent Error (MAPE)
## All states
mape <- dt2[, .(mape = 100*mean(abs(Rate - prd3)/(Rate))),
            by = set]

## By State
mape_state <- dt2[, .(mape = 100*mean(abs(Rate - prd3)/(Rate))),
                  by = list(State,
                            set)]
```

## Plot NLME
```{r,fig.height=10,fig.width=15}
p2 <- p1 +
  geom_line(data = dt2,
            aes(x = Date,
                y = prd3),
            color = "blue")
p2
```

## Table 1: Betas
```{r}
betas
```

## Table 2: MAPE, all states
```{r}
mape
```

## Table 3: MAPE, by State
```{r}
tmp <- copy(mape_state)

tmp$set <- factor(tmp$set,
                  levels = c("trn",
                             "tst",
                             "cvd"),
                  labels = c("Training",
                             "Testing",
                             "COVID-19"))

t3 <- dcast.data.table(tmp,
                       State ~ set,
                       value.var = "mape")

head(t3)

write.csv(t3,
          file = "tmp/mape_state.csv",
          row.names = FALSE)
```

# CONTINUE HERE 07/16/2023

## Save data
```{r}
dt3 <- rbindlist(dtt)
save(dt3,
     file = "tmp/dt3.RData")
```

# Excess Death
```{r}
dt3[, Deaths_prd := prd3*daysmonth*Population*10^(-5)]
```

## Over the COVID-19 months
```{r}
ed_cvd <- dt3[set == "cvd",
          .(excess_death_pct_prd = 100*sum(Deaths - Deaths_prd)/sum(Deaths_prd)),
          by = list(State,
                    Cause)]

# summary(ed_cvd)
# levels(ed_cvd$State)
# levels(ed_cvd$Cause)
# range(ed_cvd$excess_death_pct_prd)
# 
# ggplot(ed_cvd,
#        aes(x = excess_death_pct_prd)) +
#   facet_wrap(~ State,
#              scale = "free_x") +
#                geom_histogram() +
#   theme_bw()
# 
# ggplot(ed_cvd,
#        aes(x = excess_death_pct_prd)) +
#   facet_wrap(~ Cause,
#              scale = "free_x") +
#                geom_histogram() +
#   theme_bw()

t_ed_cvd <- dcast.data.table(data = ed_cvd,
                             State ~ Cause,
                             value.var = "excess_death_pct_prd")
write.csv(t_ed_cvd,
          file = "tmp/t_ed_cvd.csv",
          row.names = FALSE)
t_ed_cvd
```

## Over the Testing months
```{r}
ed_tst <- dt3[set == "tst",
              .(excess_death_pct_prd = 100*sum(Deaths - Deaths_prd)/sum(Deaths_prd)),
              by = list(State,
                        Cause)]

t_ed_tst <- dcast.data.table(data = ed_tst,
                             State ~ Cause,
                             value.var = "excess_death_pct_prd")
write.csv(t_ed_tst,
          file = "tmp/t_ed_tst.csv",
          row.names = FALSE)
t_ed_tst
```

## Plot
```{r,fig.height=8,fig.width=8}
ci_tst <- ed_tst[, .(mu = mean(excess_death_pct_prd),
                     sem = sd(excess_death_pct_prd)/sqrt(.N)),
                 by = Cause]
ci_tst[, llci := mu - 1.96*sem]
ci_tst[, ulci := mu + 1.96*sem]
ci_tst

p1 <- ggplot(ci_tst,
             aes(x = Cause,
                 y = mu,
                 ymin = llci,
                 ymax = ulci)) +
  geom_hline(yintercept = 0,
             linetype = "dashed",
             color = "red") +
  geom_errorbar(width = 0.2) +
  geom_point(shape = 21,
             size = 3,
             fill = "white") +
  geom_point(data = ed_cvd,
             aes(x = Cause,
                 y = excess_death_pct_prd,
                 group = State),
             shape = 21,
             size = 2,
             fill = "green",
             alpha = 0.5,
             position = position_dodge(0.3),
             inherit.aes = FALSE) +
  scale_y_continuous("% Predicted") +
  ggtitle("") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45,
                                   hjust = 1))

tiff(filename = paste0("tmp/tst_cvd_pctextradeath.tiff"),
     height = 8,
     width = 8,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

p1
```

## Transform data with DNAMR::trans2t
```{r}
# ed[, expp_t := trans2t(excess_death_pct_prd)$x]
# # No need? Didn't run by State or Cause!
# 
# ggplot(ed,
#        aes(x = expp_t)) +
#   facet_wrap(~ State,
#              scale = "free_x") +
#   geom_histogram() +
#   theme_bw()
```

## PCA
```{r}
# dt_pca <- dcast.data.table(data = t3,
#                            State ~ `Cause of Death`,
#                            value.var = "Training")

dt_pca <- dcast.data.table(data = ed,
                           State ~ Cause,
                           value.var = "excess_death_pct_prd")

write.csv(dt_pca,
          file = "tmp/dt_pca.csv",
          row.names = FALSE)
```

```{r}
m1 <- prcomp(dt_pca[, -1],
             center = TRUE,
             scale. = FALSE)
summary(m1)
```

```{r}
# Biplot while keep only the most important variables (Javier)----
# Select PC-s to pliot (PC1 & PC2)
choices <- 1:2
# Scores, i.e. points (df.u)
dt.scr <- data.table(m1$x[, choices])

# # Add grouping variable
dt.scr$State <- dt_pca$State
dt.scr

# Loadings, i.e. arrows (df.v)
dt.rot <- as.data.frame(m1$rotation[, choices])
dt.rot$feat <- rownames(dt.rot)
dt.rot <- data.table(dt.rot)
dt.rot
dt.load <- melt.data.table(dt.rot,
                           id.vars = "feat",
                           measure.vars = 1:2,
                           variable.name = "pc",
                           value.name = "loading")
dt.load$feat <- factor(dt.load$feat,
                     levels = unique(dt.load$feat))
dt.load

# Axis labels
u.axis.labs <- paste(colnames(dt.rot)[1:2], 
                     sprintf('(%0.1f%% explained var.)', 
                             100*m1$sdev[choices]^2/sum(m1$sdev^2)))
u.axis.labs
```

```{r,fig.width=7,fig.height=7}
scl <- 35

p1 <- ggplot(data = dt.rot,
             aes(x = PC1,
                 y = PC2)) +
  geom_point(data = dt.scr,
             # aes(fill = grp),
             shape = 21,
             size = 2,
             alpha = 0.5) +
  geom_text(data = dt.scr,
            aes(label = State),
            size = 3,
            hjust = 0.5,
            color = "red") +
  geom_segment(aes(x = 0,
                   y = 0,
                   xend = scl*PC1,
                   yend = scl*PC2),
               arrow = arrow(length = unit(1/2, 'picas')),
               color = "black",
               linewidth = 1.2) +
  geom_text(aes(x = 1.1*scl*PC1,
                y = 1.1*scl*PC2,
                label = substr(x = dt.rot$feat,
                               start = 1,
                               stop = 10)),
            size = 4,
            hjust = 0.5) +
  scale_x_continuous(u.axis.labs[1]) +
  scale_y_continuous(u.axis.labs[2]) +
  scale_fill_discrete(name = "Group") +
  ggtitle("Daily Death (% Predicted) PCA") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 20))

tiff(filename = "tmp/biplot_excess_dath_pct.tiff",
     height = 7,
     width = 7,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

p1
```

# Diabetes Death in CA Only
## Sinusoidal model NLS
```{r,fig.width=6,fig.height=4}
diab_ca <- dt3[Cause == "Diabetes mellitus (E10-E14)" &
                 State == "California"]

lm0 <- lm(Rate ~ dt + dt_sqr,
          data = diab_ca[set == "trn"])
summary(lm0)

m3 <- nls(Rate ~ b0 + b1*dt + b2*dt^2 + b3*sin(b4 + b5*dt),
          data = diab_ca[set == "trn"],
          # random = b0 + b1 + b2 + b3 + b4 + b5 ~ 1 | State,
          # fixed = b0 + b1 + b2 + b3 + b4 + b5 ~ 1,
          start = list(b0 = lm0$coefficients[1],
                       b1 = lm0$coefficients[2],
                       b2 = 0, # m0$coefficients[3],
                       b3 = 10000,
                       b4 = 1,
                       b5 = 1/60)) #  originally 0.018, 1/60=0.017
summary(m3)

diab_ca$prd4 <- predict(m3,
                        newdata = diab_ca)
```

## Plot
```{r}
p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = prd1), # prediction from lm
            color = "black",
            linetype = "dashed") +
  geom_line(data = diab_ca,
            aes(x = Date,
                y = prd3), # prediction from lme
            color = "blue") +
  geom_line(data = diab_ca,
            aes(x = Date,
                y = prd4), # prediction from nls
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```


## Fast Fourier Transform
```{r}
# x <- 1:4
# fft(x)
# xx <- fft(fft(x), inverse = TRUE)/length(x)
# 
# xx <- fft(z = diab_ca$Rate[diab_ca$set == "trn"])
# xx
```

## Spectral density of a time serie
Fourier transform decomposes functions into frequency components. Highest peak represents the largest frequency. In this case, 1/12, i.e., 12-month repeated periods.  
```{r}
xx <- spectrum(diab_ca$Rate[diab_ca$set == "trn"])

# The highest peak is around the 12-months cycle
abline(v = 1/12, 
       col = 2, # red
       lty = 2,
       lw = 2)

# Other top peaks
abline(v = 1/60,
       col = 3, # blue
       lty = 2)
abline(v = 1/6, 
       col = 4, # green
       lty = 2)

tmp <- data.table(x = 1/xx$freq,
                  y = xx$spec)
ggplot(tmp[x < 25],
       aes(x = x,
           y = y)) +
  geom_line() +
  scale_x_continuous("Months",
                     breaks = seq(from = 0, 
                                  to = 200,
                                  by = 6)) +
  scale_y_continuous("Spectral Density") +
  geom_vline(xintercept = c(6, 12),
             color = "red",
             linetype = "dashed") +
  theme_bw()
```

## Classical Seasonal Decomposition by Moving Averages
```{r,fig.width=6,fig.height=6}
xx <- ts(data = diab_ca$Rate[diab_ca$set == "trn"],
         frequency = 12)

m11 <- decompose(xx)
summary(m11)
plot(m11)
```

## Plot seasonal effect and trend
```{r}
diab_ca$seas1 <- diab_ca$trend1 <- NA
diab_ca$seas1[diab_ca$set == "trn"] <- c(m11$seasonal)
diab_ca$trend1[diab_ca$set == "trn"] <- c(m11$trend)
diab_ca[, freq1 := seas1 + trend1]

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq1),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
p1
```

```{r}
diab_ca$dt <- as.numeric(factor(diab_ca$Date))
diab_ca$dt2 <- diab_ca$dt^2

tm1 <- lm(trend1 ~ dt + dt2,
          data = diab_ca[set == "trn", ])
tm1

diab_ca$trend1_pred <- predict(tm1,
                               newdata = list(dt = diab_ca$dt,
                                              dt2 = diab_ca$dt2))
diab_ca[set == "trn",
        freq2 := seas1 + trend1_pred]

plot(trend1_pred ~ dt,
     data = diab_ca,
     type = "l")
lines(trend1 ~ dt,
     data = diab_ca,
     col = 2)

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1_pred),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq2),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
p1
```

```{r,fig.width=7,fig.height=5}
diab_ca$seas1_pred <- NA
diab_ca$seas1_pred[1:144] <- rep(diab_ca$seas1[1:12],
                                 rep = 12)
diab_ca$seas1_pred[145:150] <- diab_ca$seas1[1:6]

diab_ca[, freq3 := seas1_pred + trend1_pred]

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1_pred),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq3),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
diab_ca[set == "trn",
        freq2 := seas1 + trend1_pred]
p1
```
## MAPE
```{r}
# Seasonal predictions
diab_ca[, .(mape = 100*mean(abs(Rate - freq3)/Rate)),
        by = set]

# NLME predictions
t(t3[State == "California" &
     `Cause of Death` == "Diabetes mellitus (E10-E14)", ])
```


## Subtract seasonal effect (12-month cycle)
```{r}
# diab_ca$Rate2 <- diab_ca$Rate - c(m11$seasonal)
diab_ca[, Rate2 := Rate - seas1 - trend1_pred]
```

## Rerun spectral analysis
```{r}
xx <- spectrum(diab_ca$Rate2[diab_ca$set == "trn"])

tmp <- data.table(x = 1/xx$freq,
                  y = xx$spec)
ggplot(tmp[x < 25],
       aes(x = x,
           y = y)) +
  geom_line() +
  scale_x_continuous("Months",
                     breaks = seq(from = 0, 
                                  to = 200,
                                  by = 6)) +
  scale_y_continuous("Spectral Density") +
  geom_vline(xintercept = c(13.3),
             color = "red",
             linetype = "dashed") +
  theme_bw()
```


```{r}
xx <- ts(data = diab_ca$Rate2[diab_ca$set == "trn"],
         frequency = 13)

m12 <- decompose(xx)
summary(m12)
plot(m12)
```

In this case, 1/12 frequency seems to be the only one. 
Extend this to the training and validation sets, and calculate MAPE.  
  
# Session
```{r}
sessionInfo()
```