---
title: "COVID Excess Death"
author: Viguen Poghosyan, Davit Sargsyan
version: 2.0
output:
  html_notebook:
    highlight: tango
    toc: true
    toc_float: true
    toc_depth: 6
    number_sections: yes
    code_folding: hide
---

Date: `r date()`

Meeting 10/28/22
Javier:
1. use "decompose" function (Fourier transform)
2. find top 3 frequencies (for 12-month frequency)
3. remove 12-months signal
4. run 'decompose' again for longer frequency (24 or 36 month?)

# Setup
```{r setup}
require(data.table)
require(ggplot2)
require(nlme)
```

# Data
```{r}
dt1 <- data.table(read.csv("data/JOINT_HistRec_ETS_Results_short.csv"))
colnames(dt1)

head(dt1)
```
## Data formating
```{r}
dt2 <- dt1[, .(Date = as.Date(paste0(dt1$Month_Code,
                                     "/01"),
                              format = "%Y/%m/%d"),
               State = factor(State),
               Cause = factor(ICD.10.113.Cause.List,
                              levels = unique(ICD.10.113.Cause.List),
                              labels = gsub(x = unique(ICD.10.113.Cause.List),
                                            pattern = "#",
                                            replacement = "")),
               Population = dt1$Population,
               DeathDaily = dt1$Deaths/daysmonth,
               Rate = 10^5*(dt1$Deaths + 1)/(daysmonth*(Population + 1)))]
```

```{r}
levels(dt2$State)
levels(dt2$Cause)
```
## Remove records after 2022-07-01 and all COVID deaths
```{r}
# dt2 <- droplevels(dt2[Date < "2022-07-01" &
#                         Cause != "COVID-19 (U07.1)", ]) 
dt2 <- droplevels(dt2[Date < "2022-07-01", ]) 
```

## State populations
```{r,fig.height=10,fig.width=15}
tmp <- unique(dt2[, c("Date",
                      "State",
                      "Population")])

p1 <- ggplot(tmp,
             aes(x = Date,
                 y = Population)) +
  facet_wrap(~ State,
             scale = "free_y") +
  geom_line() +
  scale_x_continuous(breaks = seq.Date(from = min(tmp$Date),
                                       to = max(tmp$Date),
                                       by = 365.25)) +
  ggtitle("Population") +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

tiff(filename = "tmp/population.tiff",
     height = 10,
     width = 15,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

p1
```

## Daily death by State
```{r}
for (i in 1:nlevels(dt2$Cause)) {
  p1 <- ggplot(dt2[Cause == levels(dt2$Cause)[i]],
               aes(x = Date,
                   y = DeathDaily,
                   group = Cause)) +
    facet_wrap(~ State,
               scale = "free_y") +
    geom_line() +
    scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                         to = max(dt2$Date),
                                         by = 365.25)) +
    ggtitle(levels(dt2$Cause)[i]) +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90,
                                     hjust = 1))
  
  tiff(filename = paste0("tmp/",
                         levels(dt2$Cause)[i],
                         "_death_daily.tiff"),
       height = 10,
       width = 15,
       units = 'in',
       res = 300,
       compression = "lzw+p")
  print(p1)
  graphics.off()
}
```

## COVID-19 only
```{r}
  p1 <- ggplot(dt2[Cause == "COVID-19 (U07.1)" &
                     Date  >= "2020-03-01", ],
               aes(x = Date,
                   y = DeathDaily,
                   group = Cause)) +
    facet_wrap(~ State,
               scale = "free_y") +
    geom_line() +
    scale_x_continuous(breaks = seq.Date(from = as.Date("2020-03-01"),
                                         to = as.Date("2022-07-01"),
                                         by = 365.25/4)) +
    ggtitle(levels(dt2$Cause)[i]) +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90,
                                     hjust = 1))
  
  tiff(filename = "tmp/COVID-19 (U07.1)_death_daily.tiff",
       height = 10,
       width = 15,
       units = 'in',
       res = 300,
       compression = "lzw+p")
  print(p1)
  graphics.off()
```

# Remove States that never had at least 10 cerebrovascular deaths a day
```{r}
tmp <- dt2[Cause == "Cerebrovascular diseases (I60-I69)" ,
           .(maxDeath = max(DeathDaily)),
           by = State]
tmp

ndx_keep <- which(tmp$maxDeath >= 10)
tmp <- tmp[ndx_keep, ]
tmp
```

## Keep these 20 states only
```{r}
dt2 <- droplevels(dt2[State %in% tmp$State, ])
```

## Training/Testing/COVID
```{r}
dt2$set <- "trn"

dt2$set[dt2$Date >= "2019-03-01" &
          dt2$Date < "2020-03-01"] <- "tst"

dt2$set[dt2$Date >= "2020-03-01"] <- "cvd"
```

## No more zeros - recalculate the rate
```{r}
dt2[, Rate := 10^5*DeathDaily/Population]
```

## Output lists
```{r}
betas <- list()
mape <- list()
mape_state <- list()
dtt <- list()
```

## Modeling
```{r}
for (i in 1:nlevels(dt2$Cause)) {
  # Subset data
  cod <- levels(dt2$Cause)[i]
  
  tmp <- droplevels(dt2[Cause == cod,])
  
  tmp$dt <- as.numeric(tmp$Date)
  tmp$dt_sqr <- tmp$dt^2
  
  # LM
  m1 <- lm(Rate ~ (dt + dt_sqr)*State,
           data = tmp[set == "trn"])
  
  tmp$prd1 <- predict(m1,
                      newdata = tmp)
  
  # Plot
  p1 <- ggplot(tmp,
               aes(x = Date,
                   y = Rate,
                   group = Cause)) +
    facet_wrap(~ State) +
    geom_ribbon(aes(xmin = min(tmp$Date[tmp$set == "tst"]),
                    xmax = max(tmp$Date[tmp$set == "tst"]),
                    ymin = -Inf,
                    ymax = Inf),
                orientation = "y",
                fill = "green",
                alpha = 0.3) +
    geom_line(color = "gray") +
    geom_line(aes(x = Date,
                  y = prd1),
              color = "red") +
    scale_x_continuous(breaks = seq.Date(from = min(dt2$Date),
                                         to = max(dt2$Date),
                                         by = 365.25)) +
    ggtitle(cod) +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 90,
                                     hjust = 1))
  
  m3 <- try({
    # NLME
    m3 <- nlme(Rate ~ b0 + b1*dt + b2*dt^2 + b3*sin(b4 + b5*dt),
               data = tmp[set == "trn"],
               random = b0 + b1 + b2 + b3 + b4 + b5 ~ 1 | State,
               fixed = b0 + b1 + b2 + b3 + b4 + b5 ~ 1,
               start = list(fixed = c(b0 = m1$coefficients[1], # linear intercept
                                      b1 = m1$coefficients[2], # linear slope
                                      b2 = m1$coefficients[3], # square term, i.e. non-linearity of the trend
                                      b3 = 10000, # wave magnitude
                                      b4 = 1, # sin shift to the left
                                      b5 = 0.018))) # wave length(> 1 shortens, < 1 elongates))
  })
  if(class(m3)[1] != "try-error") {
    
    tmp$prd3 <- predict(m3,
                        newdata = tmp)
    
    dtt[[i]] <- tmp
    
    # Save coefficients
    betas[[i]] <- m3$coefficients$fixed
    
    # Save Mean Absolute Percent Error (MAPE)
    ## All states
    mape[[i]] <- tmp[, .(mape = 100*mean(abs(Rate - prd3)/(Rate))),
                     by = set]
    
    ## By State
    mape_state[[i]] <- tmp[, .(mape = 100*mean(abs(Rate - prd3)/(Rate))),
                           by = list(State,
                                     set)]
    # Add predicted curve
    p1 <- p1 +
      geom_line(data = tmp,
                aes(x = Date,
                    y = prd3),
                color = "blue")
  }
  
  # Save plot
  tiff(filename = paste0("tmp/",
                         cod,
                         ".tiff"),
       height = 10,
       width = 10,
       units = 'in',
       res = 300,
       compression = "lzw+p")
  print(p1)
  graphics.off()
}
```

## Table 1: Betas
```{r}
betas_t <- copy(betas)
names(betas_t) <- levels(dt2$Cause)

ndx <- lapply(X = betas_t,
              FUN = is.null)
ndx

# betas$`Malignant neoplasms (C00-C97)` <- NULL

t1 <- do.call("rbind",
              betas_t)

t1

write.csv(t1,
          file = "tmp/betas.csv",
          row.names = TRUE)
```

## Table 2: MAPE, all states
```{r}
mape_t <- copy(mape)

names(mape) <- levels(dt2$Cause)
# mape$`Malignant neoplasms (C00-C97)` <- NULL

t2 <- rbindlist(mape,
                use.names = TRUE)
t2 <- data.table(`Cause of Death` = rep(names(mape),
                                        each = 3),
                 t2)
t2$set <- factor(t2$set,
                 levels = c("trn",
                            "tst",
                            "cvd"),
                 labels = c("Training",
                            "Testing",
                            "COVID-19"))

t2 <- dcast.data.table(t2,
                       `Cause of Death` ~ set,
                       value.var = "mape")

t2

write.csv(t2,
          file = "tmp/mape.csv",
          row.names = FALSE)
```

## Table 3: MAPE, by State
```{r}
out <- list()
for (i in 1:length(mape_state)) {
  out[[i]] <- data.table(`Cause of Death` = levels(dt2$Cause)[i],
                         mape_state[[i]])
}

names(out) <- levels(dt2$Cause)

out$`COVID-19 (U07.1)` <- NULL
out$`Diseases of heart (I00-I09,I11,I13,I20-I51)` <- NULL
out$Other_rev <- NULL

t3 <- rbindlist(out)

t3$set <- factor(t3$set,
                 levels = c("trn",
                            "tst",
                            "cvd"),
                 labels = c("Training",
                            "Testing",
                            "COVID-19"))

t3 <- dcast.data.table(t3,
                       `Cause of Death` + State ~ set,
                       value.var = "mape")

head(t3)

write.csv(t3,
          file = "tmp/mape_state.csv",
          row.names = FALSE)
```

## Save data
```{r}
dt3 <- rbindlist(dtt)
save(dt3,
     file = "tmp/td3.RData")
```

# Excess Death PCA
```{r}
# dt_pca <- dcast.data.table(data = t3,
#                            State ~ `Cause of Death`,
#                            value.var = "Training")

dt_pca <- dcast.data.table(data = t3,
                           State ~ `Cause of Death`,
                           value.var = "COVID-19")

# Remove MAriland for now
dt_pca <- dt_pca[State != "Maryland",]
```

```{r}
m1 <- prcomp(dt_pca[, -1],
             center = TRUE,
             scale. = TRUE)
summary(m1)
```

```{r}
# Biplot while keep only the most important variables (Javier)----
# Select PC-s to pliot (PC1 & PC2)
choices <- 1:2
# Scores, i.e. points (df.u)
dt.scr <- data.table(m1$x[, choices])

# # Add grouping variable
dt.scr$State <- dt_pca$State
dt.scr

# Loadings, i.e. arrows (df.v)
dt.rot <- as.data.frame(m1$rotation[, choices])
dt.rot$feat <- rownames(dt.rot)
dt.rot <- data.table(dt.rot)
dt.rot
dt.load <- melt.data.table(dt.rot,
                           id.vars = "feat",
                           measure.vars = 1:2,
                           variable.name = "pc",
                           value.name = "loading")
dt.load$feat <- factor(dt.load$feat,
                     levels = unique(dt.load$feat))
dt.load

# Axis labels
u.axis.labs <- paste(colnames(dt.rot)[1:2], 
                     sprintf('(%0.1f%% explained var.)', 
                             100*m1$sdev[choices]^2/sum(m1$sdev^2)))
u.axis.labs
```

```{r,fig.width=10,fig.height=10}
scl <- 5

p1 <- ggplot(data = dt.rot,
             aes(x = PC1,
                 y = PC2)) +
  # coord_equal() +
  geom_point(data = dt.scr,
             # aes(fill = grp),
             shape = 21,
             size = 2,
             alpha = 0.5) +
    geom_text(data = dt.scr,
             aes(label = State),
             size = 3,
             hjust = 0.5,
             color = "red") +
  geom_segment(aes(x = 0,
                   y = 0,
                   xend = scl*PC1,
                   yend = scl*PC2),
               arrow = arrow(length = unit(1/2, 'picas')),
               color = "black",
               size = 1.2) +
  geom_text(aes(x = 1.1*scl*PC1,
                y = 1.1*scl*PC2,
                label = dt.rot$feat),
            size = 3,
            hjust = 0.5) +
  scale_x_continuous(u.axis.labs[1]) +
  scale_y_continuous(u.axis.labs[2]) +
  scale_fill_discrete(name = "Group") +
  ggtitle("Biplot") +
  theme_bw() +
  theme(plot.title = element_text(hjust = 0.5,
                                  size = 20))

tiff(filename = "tmp/pca_biplot_covid19.tiff",
     height = 10,
     width = 10,
     units = 'in',
     res = 300,
     compression = "lzw+p")
print(p1)
graphics.off()

p1
```

# Diabetes Death in CA Only
## Sinusoidal model NLS
```{r,fig.width=6,fig.height=4}
diab_ca <- dt3[Cause == "Diabetes mellitus (E10-E14)" &
                 State == "California"]

lm0 <- lm(Rate ~ dt + dt_sqr,
          data = diab_ca[set == "trn"])
summary(lm0)

m3 <- nls(Rate ~ b0 + b1*dt + b2*dt^2 + b3*sin(b4 + b5*dt),
          data = diab_ca[set == "trn"],
          # random = b0 + b1 + b2 + b3 + b4 + b5 ~ 1 | State,
          # fixed = b0 + b1 + b2 + b3 + b4 + b5 ~ 1,
          start = list(b0 = lm0$coefficients[1],
                       b1 = lm0$coefficients[2],
                       b2 = 0, # m0$coefficients[3],
                       b3 = 10000,
                       b4 = 1,
                       b5 = 1/60)) #  originally 0.018, 1/60=0.017
summary(m3)

diab_ca$prd4 <- predict(m3,
                        newdata = diab_ca)
```

## 
```{r}
p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = prd1), # prediction from lm
            color = "black",
            linetype = "dashed") +
  geom_line(data = diab_ca,
            aes(x = Date,
                y = prd3), # prediction from lme
            color = "blue") +
  geom_line(data = diab_ca,
            aes(x = Date,
                y = prd4), # prediction from nls
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))

p1
```


## Fast Fourier Transform
```{r}
# x <- 1:4
# fft(x)
# xx <- fft(fft(x), inverse = TRUE)/length(x)
# 
# xx <- fft(z = diab_ca$Rate[diab_ca$set == "trn"])
# xx
```

## Spectral density of a time serie
Fourier transform decomposes functions into frequency components. Highest peak represents the largest frequency. In this case, 1/12, i.e., 12-month repeated periods.  
```{r}
xx <- spectrum(diab_ca$Rate[diab_ca$set == "trn"])

# The highest peak is around the 12-months cycle
abline(v = 1/12, 
       col = 2, # red
       lty = 2,
       lw = 2)

# Other top peaks
abline(v = 1/60,
       col = 3, # blue
       lty = 2)
abline(v = 1/6, 
       col = 4, # green
       lty = 2)

tmp <- data.table(x = 1/xx$freq,
                  y = xx$spec)
ggplot(tmp[x < 25],
       aes(x = x,
           y = y)) +
  geom_line() +
  scale_x_continuous("Months",
                     breaks = seq(from = 0, 
                                  to = 200,
                                  by = 6)) +
  scale_y_continuous("Spectral Density") +
  geom_vline(xintercept = c(6, 12),
             color = "red",
             linetype = "dashed") +
  theme_bw()
```

## Classical Seasonal Decomposition by Moving Averages
```{r}
xx <- ts(data = diab_ca$Rate[diab_ca$set == "trn"],
         frequency = 12)

m11 <- decompose(xx)
summary(m11)
plot(m11)
```

## Plot seasonal effect and trend
```{r}
diab_ca$seas1 <- diab_ca$trend1 <- NA
diab_ca$seas1[diab_ca$set == "trn"] <- c(m11$seasonal)
diab_ca$trend1[diab_ca$set == "trn"] <- c(m11$trend)
diab_ca[, freq1 := seas1 + trend1]

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq1),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
p1
```

```{r}
diab_ca$dt <- as.numeric(factor(diab_ca$Date))
diab_ca$dt2 <- diab_ca$dt^2

tm1 <- lm(trend1 ~ dt + dt2,
          data = diab_ca[set == "trn", ])
tm1

diab_ca$trend1_pred <- predict(tm1,
                               newdata = list(dt = diab_ca$dt,
                                              dt2 = diab_ca$dt2))
diab_ca[set == "trn",
        freq2 := seas1 + trend1_pred]

plot(trend1_pred ~ dt,
     data = diab_ca,
     type = "l")
lines(trend1 ~ dt,
     data = diab_ca,
     col = 2)

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1_pred),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq2),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
p1
```

```{r}
diab_ca$seas1_pred <- NA
diab_ca$seas1_pred[1:144] <- rep(diab_ca$seas1[1:12],
                                 rep = 12)
diab_ca$seas1_pred[145:150] <- diab_ca$seas1[1:6]

diab_ca[, freq3 := seas1_pred + trend1_pred]

p1 <- ggplot(diab_ca,
             aes(x = Date,
                 y = Rate,
                 group = Cause)) +
  geom_ribbon(aes(xmin = min(diab_ca$Date[diab_ca$set == "tst"]),
                  xmax = max(diab_ca$Date[diab_ca$set == "tst"])),
              orientation = "y",
              fill = "green",
              alpha = 0.3) +
  geom_line(color = "gray") +
  geom_line(aes(x = Date,
                y = trend1_pred),
            color = "blue") +
  geom_line(aes(x = Date,
                y = freq3),
            color = "red") +
  scale_x_continuous(breaks = seq.Date(from = min(diab_ca$Date),
                                       to = max(diab_ca$Date),
                                       by = 365.25)) +
  ggtitle(paste(diab_ca$Cause[1],
                diab_ca$State[1],
                sep = ": ")) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90,
                                   hjust = 1))
diab_ca[set == "trn",
        freq2 := seas1 + trend1_pred]
p1
```
## MAPE
```{r}
# Seasonal predictions
diab_ca[, .(mape = 100*mean(abs(Rate - freq3)/Rate)),
        by = set]

# NLME predictions
t(t3[State == "California" &
     `Cause of Death` == "Diabetes mellitus (E10-E14)", ])
```


## Subtract seasonal effect (12-month cycle)
```{r}
# diab_ca$Rate2 <- diab_ca$Rate - c(m11$seasonal)
diab_ca[, Rate2 := Rate - seas1 - trend1_pred]
```

## Rerun spectral analysis
```{r}
xx <- spectrum(diab_ca$Rate2[diab_ca$set == "trn"])

tmp <- data.table(x = 1/xx$freq,
                  y = xx$spec)
ggplot(tmp[x < 25],
       aes(x = x,
           y = y)) +
  geom_line() +
  scale_x_continuous("Months",
                     breaks = seq(from = 0, 
                                  to = 200,
                                  by = 6)) +
  scale_y_continuous("Spectral Density") +
  geom_vline(xintercept = c(13.3),
             color = "red",
             linetype = "dashed") +
  theme_bw()
```


```{r}
xx <- ts(data = diab_ca$Rate2[diab_ca$set == "trn"],
         frequency = 13)

m12 <- decompose(xx)
summary(m12)
plot(m12)
```

In this case, 1/12 frequency seems to be the only one. 
Extend this to the training and validation sets, and calculate MAPE.  

**ToDo:**    
2. Repeat this for each of the 20 states and for each cause of death.  
3. How to make this analysis for all 20 states in a single model (similar to random effects model idea)?  
4. Use new data from Nuria (US-wide, not separated by states).    
5. F-Test residuals from the 12month and 13 month (seasonality of residuals). Other ways to compare? bootstrap?  
6. Try to get daily temperatures. What part of months are the peaks  - beginning, middle or end? are the peaks between 2 months?  
7. Using lockdown, mask and vaccination data, model residuals of the forecast (COVID years). Might be lag. Talk to Yajie.  
8. Weighted PCA? Talk to Yajie.  
9. For PCA, transform log(1+ a%) - log(1 - b%). Use DMR::trans2 function.  
  
# Session
```{r}
sessionInfo()
```